Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2014-08-15 paper.tex
More detailed info on computational resources, including XSEDE cite.
Started a discussion of the history of double degenerate modelling.

@@

This research was supported by NSF award AST-1211563.  This research
used resources of the National Energy Research Scientific Computing
Center, which is supported by the Office of Science of the
U.S. Department of Energy under Contract No. DE-AC02-05CH11231.  An
award of computer time was provided by the Innovative and Novel
Computational Impact on Theory and Experiment (INCITE) program.  This
research used resources of the Oak Ridge Leadership Computing Facility
located in the Oak Ridge National Laboratory, which is supported by
the Office of Science of the Department of Energy under Contract
DE-AC05-00OR22725.

>>>>

% History of single-degenerate model (one paragraph)

% Observational evidence that disfavors SD model (one paragraph)

% Early theoretical history of double-degenerate model (one paragraph)

% Observational evidence that favors DD model (two-three paragraphs)

The history of double degenerate theory can broadly be described as a series of three major paradigm shifts. The first attempts to model the results of the merger process came in the 1980s. \cite{nomotoiben:1985} demonstrated that if the secondary overflowed its Roche lobe and its mass accreted onto the primary near the Eddington rate, that off-center carbon ignition would occur. \cite{saionomoto:1985} tracked the evolution of the flame and found that it propagated quiescently into the center, converting the carbon-oxygen white dwarf into an oxygen-neon-magnesium white dwarf. This would then be followed by collapse into a neutron star -- a result with significantly different observational properties compared to a SN Ia. This scenario, termed accretion-induced collapse, would be avoided only if the accretion rate were well below the Eddington rate. \cite{tutukov_yungelson:1979} observed that this could happen if the mass loss from the secondary was higher than the Eddington rate and thus the accreted material formed an accretion disc, which might rain down on the primary more slowly. The main finding was that double degenerate systems would not obviously lead to Type Ia supernovae.

The first three-dimensional simulations of double degenerate systems were performed by \citet{benz:1990}, who used the smoothed particle hydrodynamics (SPH) method to simulate the merger process. They found that if the lower-mass secondary was close enough to the primary to begin mass transfer on a dynamical time scale, the secondary completely disrupted and formed a hot envelope on the primary, with an accretion disk surrounding the core and envelope. They noted that the envelope temperature was hot enough to generate carbon fusion but neglected reactions in their simulation. Later simulations of similar setups \citep{rasio_shapiro:1995,yoon:2007} corroborated this result. This validation of the prediction of \cite{tutukov_yungelson:1979} did not instill confidence in the community. \cite{mochkovitch_livio:1990} and \cite{livio:2000} observed that turbulent viscosity would be sufficiently large for angular momentum to be removed from the disk at a rate high enough to generate these troublesome accretion timescales. Based on all of this evidence, the review of \cite{hillebrandtniemeyer2000} argued that the model was only viable if the accretion-induced collapse problem could be avoided. Later work looked more carefully at the initial conditions for the simulation and reached the conclusion that mass transfer from the secondary would occur over many orbits, instead of disrupting completely over the span of a few orbits as originally suggested (ref). This may be a way to avoid the off-center carbon ignition (ref).

The most recent shift in perspective on this problem started with \cite{pakmor:2010}. This group used the SPH method to study the merger of equal-mass ($0.9\ M_\odot$) white dwarfs and found that a hotspot was generated near the surface of the primary white dwarf. They argued that this region had a temperature and density sufficient to trigger a thermonuclear detonation. They propagated this detonation throughout the system and found that it would observationally appear as a subluminous Type Ia supernova. This was the first time any simulation successfully reproduced at least some characteristics of a SN Ia. \cite{pakmor:2011} tried a few different mass combinations and found empirically that this would hold as long as the secondary was at least 80\% as massive as the primary. \cite{pakmor:2012} followed this with a study of a $1.1\ M_\odot$ and $0.9\ M_\odot$ pair, with more careful attention to the initial conditions of the simulation, and found that they could reproduce the observational characteristics of normal SNe Ia. \cite{pakmor:2013} refined this approach by adding a thin helium shell on the primary, and found that this robustly led to a detonation of the white dwarf. They argued that this framework could explain most of the observed types of SNe Ia because it allowed for both carbon-oxygen and helium white dwarfs to be the secondary, allowing for a wide range of possible outcomes.

% Discuss some other recent work (Guillochon et al. 2010, van Kerkwijk et al. 2012)

% Motivation: a series of papers to verify or disprove the prompt/violent merger hypothesis using a grid-based code

@@

This research was supported by NSF award AST-1211563.  This research
used resources of the National Energy Research Scientific Computing
Center, which is supported by the Office of Science of the
U.S. Department of Energy under Contract No. DE-AC02-05CH11231.  An
award of computer time was provided by the Innovative and Novel
Computational Impact on Theory and Experiment (INCITE) program.  This
research used resources of the Oak Ridge Leadership Computing Facility
located in the Oak Ridge National Laboratory, which is supported by
the Office of Science of the Department of Energy under Contract
DE-AC05-00OR22725. Project AST006 supported use of the ORNL/Titan resource. This work used the Extreme Science and Engineering Discovery Environment (XSEDE), which is supported by National Science Foundation grant number ACI-1053575. Project AST100037 supported use of the resources NICS/Kraken and NICS/Darter.
